---
title: "SBSW_MT_revisions"
author: "maggie weng"
date: "2/1/2024"
output: html_document
---

##Running DESeq2 differential expression analysis while correcting for batch effect in the LRT model
https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input

```{r}
#load packages
library(tidyverse)
library(DESeq2)

#load data
mt_tally <-read_csv("../mt_data_v2/mt_tally_v2.csv") #raw data from bwa
metadata_norm_v2 <-read_csv("metadata_norm_v2.csv") #metadata

#running deseq2, NOT binwise
#row names of metadata MUST equal column names of count matrix
rownames(metadata_norm_v2) <-metadata_norm_v2$sample

#first, save link btwn position and bin
bin_to_gene <-mt_tally[,c(1:4)]
#next, turn count data into matrix with positions as rownames
mt_tally_num <-mt_tally[,-c(2:4)]
mt_tally_num <-as.data.frame(mt_tally_num)
row.names(mt_tally_num) <-mt_tally_num$position
mt_tally_num$position <-NULL

dds <- DESeqDataSetFromMatrix(countData=mt_tally_num,
                              colData=metadata_norm_v2,
                              design=~extraction_batch + timecount_norm)
dds_lrt_time <- DESeq(dds, test="LRT", reduced = ~extraction_batch)
res_lrt <- results(dds_lrt_time)

save(dds_lrt_time, file="DESeq_non_binwise.Rdata")

#adjusting results for p-value cutoff of 0.05
res05 <- results(dds_lrt_time, alpha=0.05)
summary(res05)
sum(res05$padj < 0.05, na.rm=TRUE)
#test MA plot
plotMA(res05, alpha=0.05, ylim=c(-2,2))


#LFC shrinking?
#"Shrinkage of effect size (LFC estimates) is useful for visualization and ranking of genes"
#Below we specify to use the apeglm method for effect size shrinkage (Zhu, Ibrahim, and Love 2018)

resultsNames(dds_lrt_time) #we are interested in "timecount_norm" as the variable of interest
resLFC <- lfcShrink(dds_lrt_time, coef="timecount_norm", type="apeglm")

#saving results table as data table for further analysis 
res05_df <-as.data.frame(res05)
res05_df$position <-rownames(res05_df)
res05_df <-res05_df %>% left_join(., bin_to_gene, by="position")
save(res05_df, file="results_df.Rdata")

res05_df %>% filter(padj<0.05) %>% filter(product=="superoxide dismutase") #14 genes


```

#volcano plot

```{r}

res05_df$padj[is.na(res05_df$padj)] <-1 #for genes that had no pvalue, turned it into 1 so the significance would be 0 so they would appear on the plot
res05_df$significance <- "padj > 0.05"
res05_df$significance[res05_df$padj < 0.05] <- "padj < 0.05"
res05_df$significance[res05_df$padj < 0.01] <- "padj < 0.01"


#this may be too big to make one big volcano plot
#plotted one as a test
ggplot(res05_df, aes(x=log2FoldChange, y=-log10(padj), col=significance)) + geom_point() +
  scale_color_manual(values=c("blue", "red", "black")) 
```


#filter by bins to look for interesting things

201005_9_DNA_combo_bin.52: a Haloquadratum bin with high sod 
other haloquadratums with high sod:
201005_12_DNA_mat_2_bin.12
201005_12_DNA_mat_2_bin.50: Haloquadratum #3

```{r}
bin_52 <-res05_df %>% filter(bin=="201005_9_DNA_combo_bin.52") %>% filter(padj<0.05)
```


201005_12_DNA_2_bin.48: salinibacter ruber

other salinibacter bins: 201005_12_DNA_2_bin.30, 201005_12_DNA_mat_2_bin.6, 
```{r}
bin_48 <-res05_df %>% filter(bin=="201005_12_DNA_2_bin.48") %>% filter(padj<0.05)
```

Other bins of interest:
201006_1_DNA_1_bin.52, 201006_1_DNA_1_bin.17: haloarculaceae

201006_1_DNA_1_bin.34, 201007_15_DNA_mat_1_top_bin.57: haloferacaceae


Photosynthetic reaction center:
201005_10_DNA_3_bin.6: Roseovarius
201005_11_DNA_mat_1_bin.140: Rhodobacteraceae
201005_11_DNA_mat_1_bin.31: Wenzhouxiangellaceae
201007_15_DNA_mat_1_top_bin.34: Halofilum

Cyanobacteria: 
201007_15_DNA_mat_2_bottom_bin.43: Halothece
```{r}

Halothece <-res05_df %>% filter(bin=="201007_15_DNA_mat_2_bottom_bin.43") %>% filter(padj<0.05)
```


